---
title: "Lec 03 - NAs, functions, loops, and merge conflicts"
subtitle: "<br/> Statistical Programming"
author: "Fall 2021"
date: "<br/> Dr. Colin Rundel"
output:
  xaringan::moon_reader:
    css: ["slides.css"]
    lib_dir: libs
    nature:
      highlightStyle: solarized-light
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"

---
exclude: true

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
options(
  htmltools.dir.version = FALSE, # for blogdown
  width=80
)

```

---
class: middle, center

# Missing Values

---
  
## Missing Values
  
R uses `NA` to represent missing values in its data structures, what may not be obvious is that there are different `NA`s for different atomic types.

.pull-left[
```{r}
typeof(NA)
typeof(NA+1)
typeof(NA+1L)
typeof(c(NA,""))
```
]

--

.pull-right[
```{r}
typeof(NA_character_)
typeof(NA_real_)
typeof(NA_integer_)
typeof(NA_complex_)
```
]

---
  
## NA "stickiness" 
  
Because `NA`s represent missing values it makes sense that any calculation using them should also be missing.

.pull-left[
```{r}
1 + NA
1 / NA
NA * 5
```
]

.pull-right[
```{r}
sqrt(NA)
3^NA
sum(c(1, 2, 3, NA))
```
]

--

<br/>

Summarizing functions (e.g. `sum()`, `mean()`, `sd()`, etc.) will often have a `na.rm` argument which will allow you to *drop* missing values.

```{r}
sum(c(1, 2, 3, NA), na.rm = TRUE)
mean(c(1, 2, 3, NA), na.rm = TRUE)
```


---
  
## NAs are not always sticky 
  
A useful mental model for `NA`s is to consider them as a unknown value that could take any of the possible values for that type. 

--

For numbers or characters this isn't very helpful, but for a logical value we know that the value must either be `TRUE` or `FALSE` and we can use that when deciding what value to return.

--

```{r}
TRUE & NA
```

--

```{r}
FALSE & NA
```

--

```{r}
TRUE | NA
```

--

```{r}
FALSE | NA
```


---

## Conditionals and missing values

`NA`s can be problematic in some cases (particularly for control flow)

```{r error=TRUE}
1 == NA
```

--

```{r error=TRUE}
if (2 != NA)
  "Here"
```

--

```{r error=TRUE}
if (all(c(1,2,NA,4) >= 1))
  "There"
```

--

```{r error=TRUE}
if (any(c(1,2,NA,4) >= 1))
  "There"
```


---

## Testing for `NA`

To explicitly test if a value is missing it is necessary to use `is.na` (often along with `any` or `all`).

.pull-left[
```{r}
NA == NA
is.na(NA)
is.na(1)
```
]

--

.pull-right[
```{r}
is.na(c(1,2,3,NA))
any(is.na(c(1,2,3,NA)))
all(is.na(c(1,2,3,NA)))
```
]


---

## Other Special values (double)

These are defined as part of the IEEE floating point standard (not unique to R)

* `NaN` - Not a number
* `Inf` - Positive infinity
* `-Inf` - Negative infinity

.pull-left[
```{r}
pi / 0
0 / 0
1/0 + 1/0
```
]

.pull-right[
```{r}
1/0 - 1/0
NaN / NA
NaN * NA
```
]


---

## Testing for `Inf` and `NaN`

`NaN` and `Inf` don't have the same testing issues that `NA`s do, but there are still convenience functions for testing for these types of values

.pull-left[
```{r}
is.finite(Inf)
is.infinite(-Inf)
is.nan(Inf)
is.nan(-Inf)
Inf > 1
-Inf > 1
```
]

--
  
.pull-right[
```{r}
is.finite(NaN)
is.infinite(NaN)
is.nan(NaN)
is.finite(NA)
is.infinite(NA)
is.nan(NA)
```
]


---
  
## Coercion for infinity and NaN
  
First remember that `Inf`, `-Inf`, and `NaN` are doubles, however their coercion behavior is not the same as for other doubles

```{r}
as.integer(Inf)
as.integer(NaN)
```

--

.pull-left[
```{r}
as.logical(Inf)
as.logical(-Inf)
as.logical(NaN)
```
]

--

.pull-right[
```{r}
as.character(Inf)
as.character(-Inf)
as.character(NaN)
```
]

---
class: middle
count: false

# Functions

---

## Function Parts

Functions are defined by two components: the arguments (`formals`) and the code (`body`). 

Functions are assigned names like any other object in R (using `=` or `<-`)

```{r}
gcd = function(x1, y1, x2 = 0, y2 = 0) {
  R = 6371 # Earth mean radius in km
  
  # distance in km
  acos(sin(y1)*sin(y2) + cos(y1)*cos(y2) * cos(x2-x1)) * R
}
```

--

.pull-left[
```{r}
typeof(gcd)
```
]

.pull-right[
```{r}
mode(gcd)
```
]

--

<div>
.pull-left[
```{r}
formals(gcd)
```
]

.pull-right[
```{r}
body(gcd)
```
]
</div>


---

## Return values

There are two approaches to returning values from functions in R - explicit and implicit returns.

--

**Explicit** - using one or more `return` function calls

```{r}
f = function(x) {
  return(x * x)
}

f(2)
```

--

<br/>

**Implicit** - return value of the last expression is returned.

```{r}
g = function(x) {
  x * x
}

g(3)
```

---

## Invisible returns

Many functions in R make use of an invisible return value

```{r}
f = function(x) {
  print(x)
}

y = f(1)
y
```

--

```{r}
g = function(x) {
  invisible(x)
}
```

```{r}
g(2)
```

```{r}
z = g(2)
z
```



---

## Returning multiple values

If we want a function to return more than one value we can group things using atomic vectors or lists.

```{r}
f = function(x) {
  c(x, x^2, x^3)
}

f(1:2)
```

--

<br/>

```{r}
g = function(x) {
  list(x, "hello")
}

g(1:2)
```

.footnote[More on generic vectors next Tuesday]

---
class: split-50

## Argument names

When defining a function we explicitly define names for the arguments, which become variables within the scope of the function.

When calling a function we can use these names to pass arguments in an alternative order.


```{r}
f = function(x, y, z) {
  paste0("x=", x, " y=", y, " z=", z)
}
```

--

.pull-left[
```{r, error=TRUE}
f(1, 2, 3)
f(z=1, x=2, y=3)
```
]

--

.pull-right[
```{r, error=TRUE}
f(y=2, 1, 3)
f(y=2, 1, x=3)
```
]

--

```{r, error=TRUE}
f(1, 2, 3, 4)
f(1, 2, m=3)
```

---

## Argument defaults

It is also possible to give function arguments default values, so that they don't need to be provided every time the function is called.

```{r error=TRUE}
f = function(x, y=1, z=1) {
  paste0("x=", x, " y=", y, " z=", z)
}
```

--

.pull-left[
```{r error=TRUE}
f(3)
f(x=3)
```
]

--

.pull-right[
```{r error=TRUE}
f(z=3, x=2)
f(y=2, 2)
```
]

--

<br/>

```{r, error=TRUE}
f()
```

---

## Scope

R has generous scoping rules, if it can't find a variable in the functions body, it will look for it in the next higher scope, and so on.

```{r}
y = 1

f = function(x) {
  x + y
}

f(3)
```

--

```{r}
y = 1

g = function(x) {
  y = 2
  x + y
}

g(3)
y
```

---

Additionally, variables defined within a scope only persist for the duration of that scope, and do not overwrite variables at a higher scopes

```{r}
x = 1
y = 1
z = 1

f = function() {
    y = 2
    g = function() {
      z = 3
      return(x + y + z)
    }
    return(g())
}

f()

c(x,y,z)
```

---

## Exercise 1 - scope

What is the output of the following code? Explain why.

```{r eval=FALSE}
z = 1

f = function(x, y, z) {
  z = x+y

  g = function(m = x, n = y) {
    m/z + n/z
  }

  z * g()
}

f(1, 2, x = 3)
```

---

## Operators as functions

In R, operators are actually a special type of function - using backticks around the operator we can write them as functions.
 
```{r}
`+`
typeof(`+`)
```

--

.pad-top[]

```{r}
x = 4:1
x + 2
`+`(x, 2)
```


---

## Getting Help

Prefixing any function name with a `?` will open the related help file for that function.

```{r, eval=FALSE}
?`+`
?sum
```

--

For functions not in the base package, you can generally see their implementation by entering the function name without parentheses (or using the `body` function).

```{r}
lm
```

---

## Less Helpful Examples

```{r}
list

`[`

sum

`+`
```



---
class: middle
count: false

# Loops

---

## for loops

Simplest, and most common type of loop in R - given a vector iterate through the elements and evaluate the code block for each.


```{r}
is_even = function(x) {
  res = c()
  
  for(val in x) {
    res = c(res, val %% 2 == 0)
  }
  
  res
}

is_even(1:10)
```


---

## `while` loops

Repeat until the given condition is **not** met (i.e. evaluates to `FALSE`)

```{r}
make_seq = function(from = 1, to = 1, by = 1) {
  res = c(from)
  cur = from
  
  while(cur+by <= to) {
    cur = cur + by
    res = c(res, cur)
  }
  
  res
}

make_seq(1, 6)
make_seq(1, 6, 2)
```

---

## `repeat` loops

Repeat the loop until a `break` is encountered

```{r}
make_seq2 = function(from = 1, to = 1, by = 1) {
  res = c(from)
  cur = from
  
  repeat {
    cur = cur + by
    
    if (cur > to)
      break
    
    res = c(res, cur)
  }
  
  res
}

make_seq2(1, 6)
make_seq2(1, 6, 2)
```

---
class: split-50

## Special keywords - `break` and `next`

These are special actions that only work *inside* of a loop

* `break` - ends the current **loop** (inner-most)
* `next` - ends the current **iteration**

.pull-left[
```{r}
f = function(x) {
  res = c()
  
  for(i in x) {
    if (i %% 2 == 0)
      break
    
    res = c(res, i)
  }
  
  res
}

f(1:10)
f(c(1,1,1,2,2,3))
```
]

.pull-right[
```{r}
g = function(x) {
  res = c()
  
  for(i in x) {
    if (i %% 2 == 0)
      next
  
    res = c(res,i)
  }
  
  res
}

g(1:10)
g(c(1,1,1,2,2,3))
```
]


---

## Some helpful functions

Often we want to use a loop across the indexes of an object and not the elements themselves. There are several useful functions to help you do this: `:`, `length`, `seq`, `seq_along`, `seq_len`, etc.

.pull-left[
```{r}
4:7
length(4:7)
seq(4,7)
```
]

.pull-right[
```{r}
seq_along(4:7)
seq_len(length(4:7))
seq(4,7,by=2)
```
]


---

## Avoid using `1:length(x)`

A common loop construction you'll see in a lot of R code is using `1:length(x)` to generate a vector of index values for the vector `x`. 

.pull-left[ .midi[
```{r}
f = function(x) {
  for(i in 1:length(x)) {
    print(i)
  }
}

f(2:1)
f(2)
f(integer())
```
] ]

.pull-right[
```{r}
g = function(x) {
  for(i in seq_along(x)) {
    print(i)
  }
}

g(2:1)
g(2)
g(integer())
```
]

--

<div> 

.pull-left[
```{r}
1:length(integer())
```
]
.pull-right[
```{r}
seq_along(integer())
```
]

</div>



---

## Exercise 2

Below is a vector containing all prime numbers between 2 and 100:

.center[
```r
primes = c( 2,  3,  5,  7, 11, 13, 17, 19, 23, 29, 31, 37, 41, 
      43, 47, 53, 59, 61, 67, 71, 73, 79, 83, 89, 97)
```
]

If you were given the vector `x = c(3,4,12,19,23,51,61,63,78)`, write the R code necessary to print only the values of `x` that are *not* prime (without using subsetting or the `%in%` operator). 

Your code should use *nested* loops to iterate through the vector of primes and `x`.










---
class: middle center

## Merge Conflict + GitHub Actions <br/> Live Demo