---
title: "Lec 02 - Logic and types in R"
subtitle: "<br/> Statistical Programming"
author: "Fall 2021"
date: "<br/> Dr. Colin Rundel"
output:
  xaringan::moon_reader:
    css: ["slides.css"]
    lib_dir: libs
    nature:
      highlightStyle: solarized-light
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"

---
exclude: true

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
options(
  htmltools.dir.version = FALSE, # for blogdown
  width=80
)

```


---
class: middle
count: false

# In R (almost) <br/> everything is a vector

---

## Vectors

The fundamental building block of data in R are vectors (collections of related values, objects, data structures, functions, etc). 

<br/>

R has two types of vectors:

* **atomic** vectors (*vectors*)

    - homogeneous collections of the *same* type (e.g. all `true`/`false` values, all numbers, or all character strings).

* **generic** vectors (*lists*)
  
    - heterogeneous collections of *any* type of R object, even other lists <br/> 
    (meaning they can have a hierarchical/tree-like structure).

---
class: middle
count: false

# Atomic Vectors

---

## Atomic Vectors

R has six atomic vector types, we can check the type of any object in R using the `typeof()` function

`typeof()`  |  `mode()`     
:-----------|:------------
logical     |  logical    
double      |  numeric    
integer     |  numeric    
character   |  character  
complex     |  complex    
raw         |  raw        

Mode is a higher level abstraction, we will discuss this in detail a bit later.

.footnote[
There are additional types in R, e.g. generic vectors have type `list`, but more on these later. See `?typeof` for more information.
]


---

## `logical` - boolean values (`TRUE` and `FALSE`)

.pull-left[
```{r}
typeof(TRUE)
typeof(FALSE)
```
]

.pull-right[
```{r}
mode(TRUE)
mode(FALSE)
```
]

--

<br/>

R will let you use `T` and `F` as shortcuts to `TRUE` and `FALSE` is it a bad idea to use these as they are global variables that can be overwritter.

```{r}
T
T = FALSE
```

```{r}
T
```

---

## `character` - text strings

Either single or double quotes are fine, opening and closing quote must match.

.pull-left[
```{r}
typeof("hello")
typeof('world')
```
]

.pull-right[
```{r}
mode("hello")
mode('world')
```
]

--

<br/>

Quote characters can be included by escaping or using a non-matching quote.

```{r}
"abc'123"
'abc"123'
"abc\"123"
```

.footnote[RStudio's syntax highlighting is helpful here to indicate where it thinks a string begins and ends.]


---

## Numeric types

`double` - floating point values (default numerical type)

.pull-left[
```{r}
typeof(1.33)
typeof(7)
```
]

.pull-right[
```{r}
mode(1.33)
mode(7)
```
]

--

<br/>

`integer` - integer values (literals are indicated with an `L` suffix)

<div>

.pull-left[
```{r}
typeof( 7L )
typeof( 1:3 )
```
]

.pull-right[
```{r}
mode( 7L )
mode( 1:3 )
```
]

</div>

---

## Concatenation

Atomic vectors can be grown (combined) using the concatenate `c()` function.

```{r}
c(1, 2, 3)
```

--

```{r}
c("Hello", "World!")
```

--

```{r}
c(1, 1:10)
```

--

```{r}
c(1,c(2, c(3)))
```

.footnote[**Note** - atomic vectors are inherently flat.]

---
class: split-thirds

## Inspecting types

* `typeof(x)` - returns a character vector (length 1) of the *type* of object `x`.

* `mode(x)` - returns a character vector (length 1) of the *mode* of object `x`.


.pull-left[
```{r}
typeof(1)
typeof(1L)
typeof("A")
typeof(TRUE)
```
]

.pull-right[
```{r}
mode(1)
mode(1L)
mode("A")
mode(TRUE)
```
]

---

## Type Predicates

* `is.logical(x)`   - returns `TRUE` if `x` has *type* `logical`.
* `is.character(x)` - returns `TRUE` if `x` has *type* `character`.
* `is.double(x)`    - returns `TRUE` if `x` has *type* `double`.
* `is.integer(x)`   - returns `TRUE` if `x` has *type* `integer`.
* `is.numeric(x)`   - returns `TRUE` if `x` has *mode* `numeric`.

.col3_left[
```{r}
is.integer(1)
is.integer(1L)
is.integer(3:7)
```
]

.col3_mid[
```{r}
is.double(1)
is.double(1L)
is.double(3:8)
```
]

.col3_right[
```{r}
is.numeric(1)
is.numeric(1L)
is.numeric(3:7)
```
]


---

## Other useful predicates

* `is.atomic(x)` - returns `TRUE` if `x` is an *atomic vector*.
* `is.list(x)` - returns `TRUE` if `x` is a *list*.
* `is.vector(x)` - returns `TRUE` if `x` is either an *atomic vector* or *list*.

.pull-left[
```{r}
is.atomic(c(1,2,3))
is.list(c(1,2,3))
is.vector(c(1,2,3))
```
]

.pull-right[
```{r}
is.atomic(list(1,2,3))
is.list(list(1,2,3))
is.vector(list(1,2,3))
```
]


---

## Type Coercion

R is a dynamically typed language -- it will automatically convert between most types without raising warnings or errors. Keep in mind the rule that atomic vectors must always contain values of the same type.

```{r}
c(1, "Hello")
```

--

.top-pad[]

```{r}
c(FALSE, 3L)
```

--

.top-pad[]

```{r}
c(1.2, 3L)
```

---

## Operator coercion

Operators and functions will also attempt to coerce values to an appropriate type for the given operation

<div>
.pull-left[
```{r}
3.1+1L
5 + FALSE
```
]

.pull-right[
```{r}
log(1)
log(TRUE)
```
]
</div>

--

<br/><br/>

.pull-left[
```{r}
TRUE & FALSE
TRUE & 7
```
]

.pull-right[
```{r}
TRUE | FALSE
FALSE | !5
```
]


---

## Explicit Coercion

Most of the `is` functions we just saw have an `as` variant which can be used for *explicit* coercion.

.pull-left[
```{r}
as.logical(5.2)
as.character(TRUE)
as.integer(pi)
```
]

.pull-right[
```{r}
as.numeric(FALSE)
as.double("7.2")
as.double("one")
```
]

---

## Exercise 1

**Part 1**

What is the type of the following vectors? Explain why they have that type.

* `c(1, NA+1L, "C")`
* `c(1L / 0, NA)`
* `c(1:3, 5)`
* `c(3L, NaN+1L)`
* `c(NA, TRUE)`


**Part 2**

Considering only the four (common) data types, what is R's implicit type conversion hierarchy (from highest priority to lowest priority)? 

.footnote[*Hint* - think about the pairwise interactions between types.]

---
count: false
class: middle

# Conditionals & Control Flow

---

## Logical (boolean) operators

<br/><br/>

|  Operator                     |  Operation    |  Vectorized? 
|:-----------------------------:|:-------------:|:------------:
| <code>x &#124; y</code>       |  or           |   Yes        
| `x & y`                       |  and          |   Yes        
| `!x`                          |  not          |   Yes        
| <code>x &#124;&#124; y</code> |  or           |   No         
| `x && y`                      |  and          |   No         
|`xor(x, y)`                    |  exclusive or |   Yes        


---

## Vectorized?

```{r}
x = c(TRUE,FALSE,TRUE)
y = c(FALSE,TRUE,TRUE)
```



--

.pull-left[
```{r}
x | y
x & y
```
]

--

.pull-right[
```{r}
x || y
x && y
```
]

.footnote[
**Note** both `||` and `&&` only use the *first* value in the vector, all other values are ignored, there is no warning about the ignored values.
]
---

## Vectorization and math

Almost all of the basic mathematical operations (and many other functions) in R are vectorized.

.pull-left[
```{r}
c(1, 2, 3) + c(3, 2, 1)
c(1, 2, 3) / c(3, 2, 1)
```
]

--

.pull-right[
```{r}
log(c(1, 3, 0))
sin(c(1, 2, 3))
```
]

---

## Length coercion

```{r}
x = c(TRUE, FALSE, TRUE)
y = c(TRUE)
z = c(FALSE, TRUE)
```

--



.pull-left[
```{r}
x | y
x & y
```
]

.pull-right[
```{r}
y | z
y & z
```
]

<div/>

--

<br/>



```{r}
x | z
```

---

## Length coercion and math

The same length coercion rules apply for most basic mathematical operators as well

```{r}
x = c(1, 2, 3)
y = c(5, 4)
z = 10L
```

.pull-left[
```{r}
x + x
x + z
```
]

--

.pull-right[
```{r}
log(x)
y / z
```
]

--

<br/>



```{r}
x %% y
```


---

## Comparison operators

  Operator  |  Comparison                |  Vectorized?
:----------:|:--------------------------:|:----------------:
 `x < y`    |  less than                 |  Yes
 `x > y`    |  greater than              |  Yes
 `x <= y`   |  less than or equal to     |  Yes
 `x >= y`   |  greater than or equal to  |  Yes
 `x != y`   |  not equal to              |  Yes
 `x == y`   |  equal to                  |  Yes
 `x %in% y` |  contains                  |  Yes (over `x`)

.footnote[over `x` here means the returned value will have the same length as `x`]

---

## Comparisons

```{r}
x = c("A","B","C")
z = c("A")
```



.pull-left[
```{r}
x == z
x != z
x > z
```
]

--

.pull-right[
```{r}
x %in% z
z %in% x
```
]
 

---

## Conditional Control Flow

Conditional execution of code blocks is achieved via `if` statements. 

```{r}
x = c(1, 3)
```

--

```{r}
if (3 %in% x)
  print("Contains 3!")
```

--

```{r}
if (1 %in% x)
  print("Contains 1!")
```

--

```{r}
if (5 %in% x)
  print("Contains 5!")
```

--

```{r}
if (5 %in% x) {
  print("Contains 5!")
} else {
  print("Does not contain 5!")
}
```




---

## `if` is not vectorized

```{r}
x = c(1, 3)
```

--

```{r}
if (x == 1)
  print("x is 1!")
```

--

<br/>

```{r}
if (x == 3)
  print("x is 3!")
```

---

## Collapsing logical vectors

There are a couple of helper functions for collapsing a logical vector down to a single value: `any`, `all`

```{r}
x = c(3,4,1)
```

.top-pad[]

.pull-left[
```{r}
x >= 2
any(x >= 2)
all(x >= 2)
```
]

.pull-right[
```{r}
x <= 4
any(x <= 4)
all(x <= 4)
```
]

<div/>

--

<br/>

```{r}
if (any(x == 3)) 
  print("x contains 3!")
```

---

## `else if` and `else`

.pull-left[
```{r}
x = 3

if (x < 0) {
  "x is negative"
} else if (x > 0) {
  "x is positive"
} else {
  "x is zero"
}
```
]

.pull-right[
```{r}
x = 0

if (x < 0) {
  "x is negative"
} else if (x > 0) {
  "x is positive"
} else {
  "x is zero"
}
```
]

---

## `if` and `return`

If R conditional statements can return a value, the two following implementations are equivalent.

.pull-left[
```{r}
x = 5
```

```{r}
s = if (x %% 2 == 0) {
  x / 2
} else {
  3*x + 1
}
```

```{r}
s
```
]

.pull-right[
```{r}
x = 5
```

```{r}
if (x %% 2 == 0) {
  s = x / 2
} else {
  s = 3*x + 1
}
```

```{r}
s
```
]

.footnote[
Notice that conditional expressions are evaluated in the parent scope.
]

---

## Exercise 2

Take a look at the following code, without running it in R,

```r
f = function(x) {
  # Check small prime
  if (x > 10 || x < -10) {
    stop("Input too big")
  } else if (x %in% c(2, 3, 5, 7)) {
    cat("Input is prime!\n")
  } else if (x %% 2 == 0) {
    cat("Input is even!\n")
  } else if (x %% 2 == 1) {
    cat("Input is odd!\n")
  }
}
```

what do you expect the outcome will be for each of the following
```r
f(1)
f(3)
f(8)
f(-1)
f(-3)
f(1:2)
f("0")
f("3")
f("zero")
```

.footnote[More on functions next time]


---

class: middle
count: false

# Error Checking

---

## `stop` and `stopifnot`

Often we want to validate user input or function arguments - if our assumptions are not met then we often want to report the error and stop execution. 

```{r error=TRUE}
ok = FALSE
```

```{r error=TRUE}
if (!ok)
  stop("Things are not ok.")
```

```{r error=TRUE}
stopifnot(ok)
```

.footnote[*Note* - an error (like the one generated by `stop`) will prevent an RMarkdown document from compiling unless `error = TRUE` is set for that code chunk]

---

## Style choices

.pull-left[
Do stuff:
```{r eval=FALSE}
if (condition_one) {
  ##
  ## Do stuff
  ##
} else if (condition_two) {
  ##
  ## Do other stuff
  ##
} else if (condition_error) {
  stop("Condition error occured")
}
```
]

.pull-right[
Do stuff (better):
```{r eval=FALSE}
# Do stuff better
if (condition_error) {
  stop("Condition error occured")
}

if (condition_one) {
  ##
  ## Do stuff
  ##
} else if (condition_two) {
  ##
  ## Do other stuff
  ##
}
```
]

---

## Exercise 3

Write a set of conditional(s) that satisfies the following requirements,

* If `x` is greater than 3 and `y` is less than or equal to 3 then print "Hello world!"

* Otherwise if `x` is greater than 3 print "!dlrow olleH"

* If `x` is less than or equal to 3 then print "Something else ..."

* Stop execution if x is odd and y is even and report an error, don't print any of the text strings above.


Test out your code by trying various values of `x` and `y`.
