---
title: "Shiny"
subtitle: "<br/> Statistical Programming"
author: "Fall 2021"
date: "<br/> Dr. Colin Rundel"
output:
  xaringan::moon_reader:
    css: ["slides.css"]
    lib_dir: libs
    nature:
      highlightStyle: solarized-light
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:9"
---
exclude: true

```{r, message=FALSE, warning=FALSE, include=FALSE}
options(
  htmltools.dir.version = FALSE, # for blogdown
  width = 80,
  tibble.width = 80
)

knitr::opts_chunk$set(
  fig.align = "center"
)

htmltools::tagList(rmarkdown::html_dependency_font_awesome())


library(dplyr)
```

```{r include = FALSE, error=TRUE}
path = here::here("flights.sqlite")
if (file.exists(path))
  unlink(path)
```



---

## Final Projects

For the final project you will be asked to do something interesting with something that we have learned in this class. 

Anything vaguely related to statistical computing is fine

* The final project will be due on Monday December 13th by 5:00 pm.

* You can form your own team of up to 5 students to work on the project

* Project will be turned in as an R package, including a write-up (in a vignette) and all necessary data, documentation, tests, etc.

* Proposals due Monday, November 15th (requirements: team name, members, 1 paragraph description)


---
class: center, middle

# Shiny live demo - ABC
